A tool set for random number generation on GPUs in R

### 2021-12-24 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3059 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3064 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Below is an example that produces a matrix of Exponential random numbers with expectation equal to 1.
<<random exponential, eval=TRUE, cache=FALSE>>=
r_matrix <- clrng::rexp(c(2,4), rate=1, Nglobal=c(2,2), type="double")
as.matrix(r_matrix)
@
The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

 <<timecomparemonth4Cpu, cache=TRUE, eval=TRUE>>=
 #using CPU
 system.time(result_monthcpu<-stats::fisher.test(month,simulate.p.value = TRUE, B=1e6))
 result_monthcpu$p.value
 @

@@

<<TimecompareweekGpu2, eval=TRUE, cache=FALSE>>=
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_week<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
result_week$threshold
result_week$simNum
result_week$counts
result_week$p.value
@
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU. 
<<TimecompareweekGpu3, echo=FALSE, eval=TRUE, cache=TRUE>>=
result_week$cpu = as.vector(result_week$sim)
@

@@

\subsection{A summary of the results}
We summarized the comparison results in the table \ref{tab:summary} and plot the test statistics in Figure %\ref{fig4}.
<<summarycompare, echo=FALSE, eval=TRUE, cache=TRUE, fig.pos='h'>>=
library(kableExtra)
dt <- data.frame(
  var1 = c("B", "1M", "10M", "1M", "10M"),
  var2 = c('Intel 2.5ghz', 0.403804, 0.0001251, 10.74,  63.24),
  var3 = c('AMD Radeon', 0.403507, 0.0001274,   2.28,    10.82),
  var4 = c('Intel 3.7ghz', 0.4035606,0.0001202, 14.80,  92.00),
  var5 = c('NVIDIA V100', 0.403507,0.0001274,   0.32,   1.95),
  var6 = c('Data','month','week','month','week')
)

@@

<<fighistMonth, eval=TRUE, echo=FALSE, dev='pdf', fig.cap = "Approximate sampling distributions of the test statistics from the two examples Month and Week. The test statistic values of the observed tables are marked with a blue line on each plot.", fig.subcap = c('Month data', 'week data'), out.width="0.47\\textwidth", margins=1>>=
hist(month_stats, xlab="test statistic",#TeX('-$\\sum(\\log(n_{ij}!))$'),
     breaks=40, ylab="proportion", main="", prob=TRUE)
abline(v = result_month$threshold, col = "blue", lwd = 1.5)

@@

\section{An application of normal RNG: Gaussian surface simulation} 
Given a parameter space $X$, a random field $U$ on $X$ is a collection of random variables $\{U_x :x\in X\}$. A Gaussian random field is a random field with the property that for any positive integer $n$ and any set of locations $x_1,\dots,x_n \in X$, the joint distribution of $U=(U_{x_1},\dots,U_{x_n})$ is multivariate Gaussian. The expectation of $U$ and covariance function $\Sigma$ completely determine the distribution of $U$, where 
\begin{gather*} 
U = [U(x_1), \ldots, U(x_n)]^\top \sim \text{MVN}(0, \Sigma) \\
\Sigma_{ij} = \cov[ U(x_i),U(x_j) ] = \sigma^2 \frac{2^{\kappa-1}}{\Gamma(\kappa)} \left(\sqrt{8\kappa} \frac{|x_i = x_j|}{\phi}\right)^\kappa  K_\kappa\left(\sqrt{8\kappa}  \frac{|x_i = x_j|}{\phi}\right).
\end{gather*}
A common choice for the covariance function $\Sigma$ is the Mat\'ern covariance \citep{matern1960spatial}. The Mat\'ern covariance between $U(x_i)$ and $U(x_j)$ takes the above form,
%$d =\|x-x'\|$ is the Euclidean distance between two spatial points, 
where $\sigma^2$ the variance of the random field $U$.
$\Gamma(\cdot)$ is the standard gamma function.  
$\phi$ is the range parameter or scale parameter with the dimension of distance, it controls the rate of decay of the correlation as distance increases. $K_\kappa(\cdot)$ is the modified Bessel function of the second kind with order $\kappa$, $\kappa$ is the shape parameter which determines the smoothness of $U(x)$, specifically, $U(x)$ is $m$ times mean-square differentiable if and only if $\kappa > m$. %(For $\kappa=0.5$, the Mat\'ern covariance function reduces to the exponential covariance function $\sigma^2*\exp(-d/\phi)$, when $\kappa \rightarrow \infty$, $\Sigma(d) \rightarrow \sigma^2*\exp{-(\|d\|/\phi)^2}$, which is called the Gaussian covariance.)  
There are several alternative parameterisations of Mat\'ern covariance functions in literatures  \citep[see][]{haskard2007anisotropic}. In \pkg{clrng}, we use the above form for computing Mat\'ern covariance.

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on 7 popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except matrix decomposition method, are approximations and has specific requirements on the type of grid or covariance functions. The matrix decomposition method is exact, it works for all covariance functions and can generate random field on all types of grids. Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations; the \pkg{RandomFields} \citep{RandomFields2015,RandomFields2020} package can use different methods for simulation of Gaussian fields, among which the modified circulant embedding method \citep{Dietrich1997FastAE} for covariance matrix decomposition is also an exact method, however, it works only for isotropic Gaussian fields and on rectangular grids. \pkg{clrng} does simulation of exact Gaussian fields on the GPU as it relies on the matrix decomposition method. The implementation of this method is as follows.

Suppose $U=(U_1, U_2, \dots, U_n)$ is a Gaussian random field with mean $\mu$ and covariance matrix $\Sigma$, without loss of generality, we assume mean value zero $(\mu = 0)$. Since covariance matrix $\Sigma$ is symmetric and positive-definite, we can take Cholesky decomposition of $\Sigma$, then we can simulate random samples from $U$ using Algorithm \ref{algorithm2},

\begin{algorithm}[H]
\SetAlgoLined
 1, Calculate the covariance matrix $\Sigma$ between locations\;
 2, Compute the Cholesky decomposition of $\Sigma = L \cdot D \cdot L^\top$, $L$ is the lower unit triangular matrix, $D$ is a diagonal matrix\;
 3, Generate on GPU a random matrix $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix\;
 4, Compute the random samples from $U$ in batches using $U = L \cdot D^{\frac{1}{2}} \cdot Z$ \;
 \caption{Gaussian random fields simulation using covariance matrix decomposition method.}
 \label{algorithm2}
\end{algorithm}

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The matrix decomposition method is straightforward to implement. However, when the number of locations $n$ is large, the cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and the cost of matrix-vector multiplication $L*Z$ to generate the random field $U$ is $\mathcal{O}(n^2)$ \citep{LiuandLi2019}.  Our another \proglang{R} package \pkg{gpuBatchMatrix} is able to compute batches of Mat\'ern covariance matrices in parallel, and do Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU. The following shows an example, in which we simulate on GPU 10 Gaussian random fields of mat\'ern covariance at one time with 5 sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

@@

<<swissRainBoundary, echo=TRUE>>=
data("swissRain", package="geostatsp")
myRaster = geostatsp::squareRaster(swissBorder, 80)
myRaster
@
Step 2, create 4 parameter sets as a small example. 
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
params = 
rbind(c(shape=1.25, range=50*1000, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=60*1000, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.6, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7),
c(shape=3, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7)
)
@
<<show parameterBatch,eval=TRUE>>=
params
@
Step 3, compute the Mat\'ern covariance matrices using \fct{gpuBatchMatrix::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4080 \times 4080$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
maternCov = gpuBatchMatrix::maternBatch(
  params, myRaster,          
  Nglobal=c(128,64), Nlocal=c(16,4))
dim(maternCov)
@
Step 4, perform Cholesky decomposition of \code{maternCov}. The first argument in \fct{gpuBatchMatrix::cholBatch} specifies the object to take Cholesky decomposition. Computed unit lower triangular matrices $L_i$'s are stacked by row and stored in \code{maternCov}. The diagonal values of each $D_i$ are returned and stored in each row of \code{diagMat}. For example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = gpuBatchMatrix::cholBatch(maternCov, Nglobal= c(128, 8), Nlocal= c(32, 8))
@
\begin{gather}
\begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
\end{bmatrix}
 \rightarrow
 \begin{bmatrix}
  L_{1} \\ L_{2} \\L_{3} \\ \vdots
  \end{bmatrix} \text{and}
  \begin{bmatrix}
  D_{1} \\ D_{2} \\D_{3} \\ \vdots
  \end{bmatrix}
\end{gather}
Step 5, create some streams on GPU to generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{clrng::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- createStreamsGpu(n=128*64)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = clrng::rnorm(
                c(nrow(maternCov),2), streams=streamsGpu, 
                Nglobal=c(128,64),
                type = theType)
@
Step 6, compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration for the matrix operation that \fct{multiplyLowerDiagonalBatch} does.  \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = gpuBatchMatrix::multiplyLowerDiagonalBatch(maternCov, 
  diagMat, zmatGpu,
  diagIsOne = TRUE,   
  transformD = "sqrt", 
  Nglobal=c(128, 64, 2), 
  Nlocal= c(8, 2, 1), 
  NlocalCache=1000)
@
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} \\
   \vdots  &   \vdots
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 realizations.
<<r plotSimSetup, echo=TRUE, eval=TRUE, cache=FALSE>>=
simRaster = raster::brick(myRaster, nl = ncol(simMat)*nrow(params))
values(simRaster) = as.vector(as.matrix(simMat))
@

@@

\section{Discussion}
The package \pkg{clrng} has been created to make GPU-generated uniform and normal random numbers accessible for \proglang{R} users, it enables reproducible research in simulations by setting seeds in streams on GPU. We further applied the GPU-generated random numbers in suitable statistical simulations, such as the Monte Carlo simulation for Fisher’s exact test, and (exact) Gaussian spatial surfaces simulation, for which we are able to calculate quantiles for the normal distribution, compute mat\'ern covariance matrices, do Cholesky decomposition and matrix multiplication in batches on GPU. Most of these functions uses local memory on GPU. Comparison of performance between using \pkg{clrng} and using classic \proglang{R} on CPU for some real data examples has demonstrated significant improvement in execution time.

@@

\pkg{clrng} is limited by the number and type of \proglang{OpenCL} RNGs used and the features of the \pkg{gpuR} package, as it depends upon the \pkg{gpuR} package, if developers wants to use our package to do something that's not supported in \pkg{gpuR}, for example ``sparse'' class objects, they would have to write \proglang{OpenCL} code to implement it.

>>>>

Below is an example that produces a $2 \times 4$ matrix of Exponential random numbers with expectation equal to 1.
<<random exponential, eval=TRUE, cache=FALSE>>=
r_matrix <- clrng::rexp(c(2,4), rate=1, Nglobal=c(2,2), type="double")
as.matrix(r_matrix)
@
The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

 <<timecomparemonth4Cpu, cache=TRUE, eval=TRUE>>=
 #using CPU
 system.time(result_monthcpu<-stats::fisher.test(month,simulate.p.value = TRUE, B=1015808))
 result_monthcpu$p.value
 @

@@

<<TimecompareweekGpu2, eval=TRUE, cache=FALSE>>=
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_week<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
result_week$threshold
result_week$simNum
result_week$counts
result_week$p.value
@
The ``week'' table has a much smaller p-value: around 0.000125, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 92.45 seconds, while \fct{fisher.sim} takes about 1.95 seconds, the GPU elapsed time is decreased to about 2.10\% of the time taken on CPU. 
<<TimecompareweekGpu3, echo=FALSE, eval=TRUE, cache=TRUE>>=
result_week$cpu = as.vector(result_week$sim)
@

@@

\subsection{A summary of the results}
We summarized the comparison results in Table \ref{tab:summary} and plot the test statistics in Figure \ref{fig4}.
<<summarycompare, echo=FALSE, eval=TRUE, cache=TRUE, fig.pos='h'>>=
library(kableExtra)
dt <- data.frame(
  var1 = c("B", "1M", "10M", "1M", "10M"),
  var2 = c('Intel 2.5ghz', 0.403804, 0.0001251, 10.74,  63.24),
  var3 = c('AMD Radeon', 0.403507, 0.0001274,   2.28,    10.82),
  var4 = c('Intel 3.7ghz', 0.4035606,0.0001202, 14.80,  92.00),
  var5 = c('NVIDIA V100', 0.403507,0.0001274,   0.32,   1.95),
  var6 = c('Data','month','week','month','week')
)

@@

<<fighistMonth, eval=TRUE, echo=FALSE, dev='pdf', fig.cap = "\\label{fig4}Approximate sampling distributions of the test statistics from the two examples Month and Week. The test statistic values of the observed tables are marked with a blue vertical line on each plot.", fig.subcap = c('Month data', 'week data'), out.width="0.47\\textwidth", margins=1>>=
hist(month_stats, xlab="test statistic",#TeX('-$\\sum(\\log(n_{ij}!))$'),
     breaks=40, ylab="proportion", main="", prob=TRUE)
abline(v = result_month$threshold, col = "blue", lwd = 1.5)

@@

\section{An application of normal RNG: Gaussian surface simulation} 
Given a parameter space $X$, a random field $U$ on $X$ is a collection of random variables $\{U_x :x\in X\}$. A Gaussian random field is a random field with the property that for any positive integer $n$ and any set of locations $x_1,\dots,x_n \in X$, the joint distribution of $U=(U_{x_1},\dots,U_{x_n})$ is multivariate Gaussian. The expectation of $U$ and covariance function $\Sigma$ completely determine the distribution of $U$, where 
\begin{gather*} 
U = [U(x_1), \ldots, U(x_n)]^\top \sim \text{MVN}(0, \Sigma), \\
\Sigma_{ij} = \cov[ U(x_i),U(x_j) ] = \sigma^2 \frac{2^{\kappa-1}}{\Gamma(\kappa)} \left(\sqrt{8\kappa} \frac{|x_i = x_j|}{\phi}\right)^\kappa  K_\kappa\left(\sqrt{8\kappa}  \frac{|x_i = x_j|}{\phi}\right),
\end{gather*}
where \text{MVN} standards for the multivariate normal distribution, and $\Sigma$ is the covariance matrix. A common choice for $\Sigma$ is the Mat\'ern covariance \citep{matern1960spatial}. The Mat\'ern covariance between $U(x_i)$ and $U(x_j)$ takes the above form,
%$d =\|x-x'\|$ is the Euclidean distance between two spatial points, 
where $\sigma^2$ the variance of the random field $U$.
$\Gamma(\cdot)$ is the standard gamma function.  
$\phi$ is the range parameter or scale parameter with the dimension of distance, it controls the rate of decay of the correlation as distance increases. $K_\kappa(\cdot)$ is the modified Bessel function of the second kind with order $\kappa$, $\kappa$ is the shape parameter which determines the smoothness of $U(x)$, specifically, $U(x)$ is $m$ times mean-square differentiable if and only if $\kappa > m$. %(For $\kappa=0.5$, the Mat\'ern covariance function reduces to the exponential covariance function $\sigma^2*\exp(-d/\phi)$, when $\kappa \rightarrow \infty$, $\Sigma(d) \rightarrow \sigma^2*\exp{-(\|d\|/\phi)^2}$, which is called the Gaussian covariance.)  
There are several alternative parameterisations of Mat\'ern covariance functions in literature  \citep[see][]{haskard2007anisotropic}. In \pkg{clrng}, we use the above form for computing Mat\'ern covariance.

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on seven popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except matrix decomposition method, are approximations and has specific requirements on the type of grid or covariance functions. The matrix decomposition method is exact, it works for all covariance functions and can generate random field on all types of grids. Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations; the \pkg{RandomFields} \citep{RandomFields2015,RandomFields2020} package can use different methods for simulation of Gaussian fields, among which the modified circulant embedding method \citep{Dietrich1997FastAE} for covariance matrix decomposition is also an exact method, however, it works only for isotropic Gaussian fields and on rectangular grids. \pkg{clrng} does simulation of exact Gaussian fields on the GPU as it relies on the matrix decomposition method. The implementation of this method is as follows.

Suppose $U=(U_1, U_2, \dots, U_n)^\top$ is a Gaussian random field with mean $\mu$ and covariance matrix $\Sigma$, without loss of generality, we assume mean value zero $(\mu = 0)$. Since covariance matrix $\Sigma$ is symmetric and positive-definite, we can take Cholesky decomposition of $\Sigma$, then we can simulate random samples from $U$ using Algorithm \ref{algorithm2},

\begin{algorithm}[H]
\SetAlgoLined
 1, Calculate the covariance matrix $\Sigma$ between locations\;
 2, Compute the Cholesky decomposition of $\Sigma = L \cdot D \cdot L^\top$, where $L$ is a lower unit triangular matrix, and $D$ is a diagonal matrix\;
 3, Generate on GPU a random matrix $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix\;
 4, Compute the random samples from $U$ in batches using $U = L \cdot D^{\frac{1}{2}} \cdot Z$ \;
 \caption{Gaussian random fields simulation using covariance matrix decomposition method.}
 \label{algorithm2}
\end{algorithm}

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The matrix decomposition method is straightforward to implement. However, when the number of locations $n$ is large, the cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and the cost of matrix-vector multiplication $L*Z$ to generate the random field $U$ is $\mathcal{O}(n^2)$ \citep{LiuandLi2019}.  Our other \proglang{R} package \pkg{gpuBatchMatrix} is able to compute batches of Mat\'ern covariance matrices in parallel, and do Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU. The following shows an example, in which we simulate on GPU eight Gaussian random fields of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

@@

<<swissRainBoundary, echo=TRUE>>=
data("swissRain", package="geostatsp")
myRaster = geostatsp::squareRaster(swissBorder, 80)
myRaster
@
Step 2, create 4 parameter sets as a small example. 
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
params = 
rbind(c(shape=1.25, range=50*1000, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=60*1000, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.6, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7),
c(shape=3, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7)
)
@
<<show parameterBatch,eval=TRUE>>=
params
@
Step 3, compute the Mat\'ern covariance matrices using \fct{gpuBatchMatrix::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4080 \times 4080$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
maternCov = gpuBatchMatrix::maternBatch(
  params, myRaster,          
  Nglobal=c(128,64), Nlocal=c(16,4))
dim(maternCov)
@
Step 4, perform Cholesky decomposition of \code{maternCov}. The first argument in \fct{cholBatch} specifies the object to take Cholesky decomposition. Computed unit lower triangular matrices $L_i$'s are stacked by row and stored in \code{maternCov}. The diagonal values of each $D_i$ are returned and stored in each row of \code{diagMat}. For example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = gpuBatchMatrix::cholBatch(maternCov, Nglobal= c(128, 8), Nlocal= c(32, 8))
@
\begin{gather}
\begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
\end{bmatrix}
 \rightarrow
 \begin{bmatrix}
  L_{1} \\ L_{2} \\L_{3} \\ \vdots
  \end{bmatrix} \text{and}
  \begin{bmatrix}
  D_{1} \\ D_{2} \\D_{3} \\ \vdots
  \end{bmatrix}
\end{gather}
Step 5, create some streams on GPU to generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{clrng::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- createStreamsGpu(n=128*64)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = clrng::rnorm(
                c(nrow(maternCov),2), streams=streamsGpu, 
                Nglobal=c(128,64),
                type = theType)
@
Step 6, compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration for the matrix operation that \fct{multiplyLowerDiagonalBatch} does.  \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = gpuBatchMatrix::multiplyLowerDiagonalBatch(maternCov, 
  diagMat, zmatGpu,
  diagIsOne = TRUE,   
  transformD = "sqrt", 
  Nglobal=c(128, 64, 2), 
  Nlocal= c(8, 2, 1), 
  NlocalCache=1000)
@
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} \\
   \vdots  &   \vdots
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 realizations.
<<r plotSimSetup, echo=TRUE, eval=TRUE, cache=FALSE>>=
simRaster = raster::brick(myRaster, nl = ncol(simMat)*nrow(params))
values(simRaster) = as.vector(as.matrix(simMat))
@

@@

\section{Discussion}
The package \pkg{clrng} has been created to make GPU-generated uniform and normal random numbers accessible for \proglang{R} users, it enables reproducible research in simulations by setting seeds in streams on GPU. We further applied the GPU-generated random numbers in suitable statistical simulations, such as the Monte Carlo simulation for Fisher’s exact test, and (exact) Gaussian spatial surfaces simulation, for which we are able to calculate quantiles for the normal distribution, compute Mat\'ern covariance matrices, do Cholesky decomposition and matrix multiplication in batches on GPU. Most of these functions uses local memory on GPU. Comparison of performance between using \pkg{clrng} and using classic \proglang{R} on CPU for some real data examples has demonstrated significant improvement in execution time.

@@

\pkg{clrng} is limited by the number and type of \proglang{OpenCL} RNGs used and the features of the \pkg{gpuR} package, as it depends upon the \pkg{gpuR} package. If developers wants to use our package to do something that's not supported in \pkg{gpuR}, for example ``sparse'' class objects, they would have to write \proglang{OpenCL} code to implement it.
